ChemKED: a human- and machine-readable data standard for chemical kinetics experiments

### 2017-07-25 pyked.tex
Minor updates to wording and text

%====================================================================
\begin{abstract} % not to exceed 200 words
Fundamental experimental measurements of quantities including ignition delay times,
laminar flame speeds, and species profiles serve important roles in understanding
fuel chemistry and validating chemical kinetic models. However, despite both the
importance and abundance of such information in the literature, the community lacks
a widely adopted standard format for this data. This impedes both sharing and wide
use by the community. Here we introduce a new chemical kinetics
experimental data format, ChemKED, and the related Python-based package for creating
and validating ChemKED-formatted files called PyKED. We also review past and
related efforts, and motivate the need for a new solution. ChemKED currently supports
the representation of autoignition delay time measurements from shock tubes and rapid
compression machines.
ChemKED-formatted files contain all of the information needed to simulate experimental
data points, including the uncertainty of the data. ChemKED is based on the YAML data
serialization language, and is intended as a human- and machine-readable standard for
easy creation and automated use.
Development of ChemKED and PyKED occurs openly on GitHub under the
BSD 3-clause license, and contributions from the community are welcome. Plans for
future development include support for experimental data from laminar flame,
jet stirred reactor, and speciation measurements.
\end{abstract}

@@

The combustion community has not yet widely adopted a standard format for
experimental data, but some researchers and groups have proposed solutions to
this problem. Most notably, Frenklach developed the PrIMe (\textbf{Pr}ocess
\textbf{I}nformatics \textbf{M}od\textbf{e}l) data
format~\autocite{Frenklach:2007bm,You:2011hy} and an associated online database,
the PrIMe Data Warehouse.\footnote{\url{http://primekinetics.org}}
PrIMe files encode fundamental
combustion experimental data, along with kinetic models and calculated
quantities, using the Extensible Markup Language (XML) standard.

PrIMe~\autocite{Frenklach:2007bm,You:2011hy} has a number of features that make
it a strong standardization format, but it suffers from several flaws that
prevent wide adoption. First, the PrIMe standard does not require or support all
the information needed to simulate an experiment, including a machine-readable
definition of ignition or a standard way to express detailed facility-specific
effects necessary to properly simulate certain experiments. Second, the XML
format is intended to be a markup language rather than a data format, and
its lack of human readibility presents a barrier to creating and working with
database files. Furthermore, the
focus on using internal identifiers for bibliographic references and
species\slash reactions, rather than standard identifiers (e.g., DOIs for
scholarly products, InCHI or SMILES for species), limits usability outside
the PrIMe ecosystem. Finally, the closed and opaque natures of the PrIMe standard
and associated Data Warehouse limit contributions and community development of
the PrIMe standard and compatible data files and tools.

More recently, Varga et al.~\autocite{Varga2015a,Varga2015b} developed the
ReSpecTh standard, which builds on the PrIMe format. ReSpecTh adds important
features that make files better standalone representations of experimental
data---i.e., more informative by themselves rather than in concert with a larger
system. For example, where PrIMe uses internal bibliographic references,
ReSpecTh adds a field for typical bibliographic data. The ReSpecTh standard also
provides machine-readable formats for describing ignition experiments, including
a field for the definition of autoignition and the ability to specify
facility-specific effects. However, ReSpecTh experimental files are another
XML-based format, and, as such, suffer from the same usability issues as PrIMe.
Moreover, the standard remains closed, requiring potential users to register
with the standard's authors to access the specification and the growing database.

Nonetheless, the usefulness of databases of fundamental combustion experiments
has been shown to be quite useful. For instance, Olm and colleagues used their
ReSpecTh-based database to quantify the performance of literature
hydrogen~\autocite{Olm:2014gn} and syngas~\autocite{Olm:2015ch} kinetic models,
and more recently develop improved models for hydrogen\slash
syngas~\autocite{Varga:2016gj} and ethanol~\autocite{Olm:2016et}. In all four
cases, they converted numerous experimental datasets from the literature into
the ReSpecTh standard.

@@

The \pybox|ChemKED| class defines three instance methods:
\pybox|get_dataframe()|, \pybox|write_file()|, and \pybox|convert_to_ReSpecTh()|.
The \pybox|get_dataframe()| method provides an instance of a
Pandas \pybox|DataFrame|~\autocite{pandas} that contains the information in the
list of \pybox|DataPoint|s. The user can specify the columns included in the
\pybox|DataFrame| by passing a list of column names to the
\pybox|output_columns| argument of the \pybox|get_dataframe()| method. The
possible columns are not case-sensitive and are:
%
\noindent\begin{itemize*}
    \item \pybox|'Temperature'|
    \item \pybox|'Pressure'|
    \item \pybox|'Ignition Delay'|
    \item \pybox|'Composition'|
    \item \pybox|'Equivalence Ratio'|
    \item \pybox|'Reference'|
    \item \pybox|'Apparatus'|
    \item \pybox|'Experiment Type'|
    \item \pybox|'File Author'|
    \item \pybox|'File Version'|
    \item \pybox|'ChemKED Version'|
\end{itemize*}

@@

Only the first author is printed when \pybox|Reference| or
\pybox|Reference:Authors| is selected because the whole author list may be
quite long. The \pybox|write_file()| method writes another \ck{} YAML file based on
the internal attributes, while \pybox|convert_to_ReSpecTh()| converts the \ck{}
to a ReSpecTh XML file. In the latter case, some information may be lost, as \ck{}
files both encode more details and specify more granularity of some information than
ReSpecTh files.

@@

# Load the ChemKED file and retrieve the first element of the
# datapoints list, which is an instance of the DataPoint class
ck = ChemKED('testfile_rcm.yaml')
dp = ck.datapoints[0]
\end{pythonbox}
%
The initial temperature, pressure, and mixture composition can be read from the
instance of the \pybox|DataPoint| class. \pk{} uses Pint \pybox|Quantities| to
store values with units, while Cantera expects a floating-point value in SI
units as input. Therefore, we use the built-in capabilities of Pint to convert
the units from those specified in the \ck{} file to SI units:
%
\begin{pythonbox}
T_initial = dp.temperature.to('K').magnitude
P_initial = dp.pressure.to('Pa').magnitude
X_initial = dp.get_cantera_mole_fraction()

>>>>

%====================================================================
\begin{abstract} % not to exceed 200 words
Fundamental experimental measurements of quantities such as ignition delay times,
laminar flame speeds, and species profiles (among others) serve important roles in understanding
fuel chemistry and validating chemical kinetic models. However, despite both the
importance and abundance of such information in the literature, the community lacks
a widely adopted standard format for this data. This impedes both sharing and wide
use by the community. Here we introduce a new chemical kinetics
experimental data format, ChemKED, and the related Python-based package for creating
and validating ChemKED-formatted files called PyKED. We also review past and
related efforts, and motivate the need for a new solution. ChemKED currently supports
the representation of autoignition delay time measurements from shock tubes and rapid
compression machines.
ChemKED-formatted files contain all of the information needed to simulate experimental
data points, including the uncertainty of the data. ChemKED is based on the YAML data
serialization language, and is intended as a human- and machine-readable standard for
easy creation and automated use.
Development of ChemKED and PyKED occurs openly on GitHub under the
BSD 3-clause license, and contributions from the community are welcome. Plans for
future development include support for experimental data from laminar flame,
jet stirred reactor, and speciation measurements.
\end{abstract}

@@

The combustion community has not yet widely adopted a standard format for
experimental data, but some researchers and groups have proposed solutions to
this problem. Most notably, Frenklach developed the PrIMe (\textbf{Pr}ocess
\textbf{I}nformatics \textbf{M}od\textbf{e}l) data
format~\autocite{Frenklach:2007bm,You:2011hy} and an associated online database,
the PrIMe Data Warehouse.\footnote{\url{http://primekinetics.org}}
PrIMe files encode fundamental
combustion experimental data, along with kinetic models and calculated
quantities, using the eXtensible Markup Language (XML) standard.

PrIMe~\autocite{Frenklach:2007bm,You:2011hy} has a number of features that make
it a strong standardization format, but it suffers from several flaws that
prevent wide adoption. First, the PrIMe standard does not require or support all
the information needed to simulate an experiment, including a machine-readable
definition of ignition or a standard way to express detailed facility-specific
effects necessary to properly simulate certain experiments. Second, the XML
format is intended to be a machine-readable markup language rather than a data format, and
its lack of human readibility presents a barrier to creating and working with
database files. Furthermore, the
focus on using internal identifiers for bibliographic references and
species\slash reactions, rather than standard identifiers (e.g., DOIs for
scholarly products, InCHI or SMILES for species), limits usability outside
the PrIMe ecosystem. Finally, the closed and opaque natures of the PrIMe standard
and associated Data Warehouse limit contributions and community development of
the PrIMe standard and compatible data files and tools.

More recently, Varga and coworkers~\autocite{Varga2015a,Varga2015b} developed the
ReSpecTh standard, which builds on the PrIMe format. ReSpecTh adds important
features that make files better standalone representations of experimental
data---i.e., more informative by themselves rather than in concert with a larger
system. For example, where PrIMe uses internal bibliographic references,
ReSpecTh adds a field for typical bibliographic data. The ReSpecTh standard also
provides machine-readable formats for describing ignition experiments, including
a field for the definition of autoignition and the ability to specify
facility-specific effects. However, ReSpecTh experimental files are another
XML-based format, and, as such, suffer from the same usability issues as PrIMe.
Moreover, the standard remains closed, requiring potential users to register
with the standard's authors to access the specification and the growing database.

Nonetheless, databases of fundamental combustion experiments
have been shown to be quite useful. For instance, Olm and colleagues used their
ReSpecTh-based database to quantify the performance of literature
hydrogen~\autocite{Olm:2014gn} and syngas~\autocite{Olm:2015ch} kinetic models,
and more recently develop improved models for hydrogen\slash
syngas~\autocite{Varga:2016gj} and ethanol~\autocite{Olm:2016et}. In all four
cases, they converted numerous experimental datasets from the literature into
the ReSpecTh standard.

@@

The \pybox|ChemKED| class defines three instance methods:
\pybox|get_dataframe()|, \pybox|write_file()|, and \pybox|convert_to_ReSpecTh()|.
The \pybox|get_dataframe()| method returns an instance of a
Pandas \pybox|DataFrame|~\autocite{pandas} that contains the information in the
list of \pybox|DataPoint|s. The user can specify the columns included in the
\pybox|DataFrame| by passing a list of column names to the
\pybox|output_columns| argument of the \pybox|get_dataframe()| method. The
possible columns are not case-sensitive and are:
%
\noindent\begin{itemize*}
    \item \pybox|'Temperature'|
    \item \pybox|'Pressure'|
    \item \pybox|'Ignition Delay'|
    \item \pybox|'Composition'|
    \item \pybox|'Equivalence Ratio'|
    \item \pybox|'Reference'|
    \item \pybox|'Apparatus'|
    \item \pybox|'Experiment Type'|
    \item \pybox|'File Author'|
    \item \pybox|'File Version'|
    \item \pybox|'ChemKED Version'|
\end{itemize*}

@@

Only the first author is printed when \pybox|Reference| or
\pybox|Reference:Authors| is selected because the whole author list may be
quite long.

The \pybox|write_file()| method writes a new \ck{} YAML file based on
the class' instance attributes, while \pybox|convert_to_ReSpecTh()| converts the \pybox|ChemKED|
instance to a ReSpecTh XML file. In the latter case, some information may be lost, as \ck{}
files both encode more details and specify more granularity of some information than
ReSpecTh files.

@@

# Load the ChemKED file and retrieve the first element of the
# datapoints list, which is an instance of the DataPoint class
ck = ChemKED('testfile_rcm.yaml')
dp = ck.datapoints[0]
\end{pythonbox}
%
The initial temperature, pressure, and mixture composition can be read from the
instance of the \pybox|DataPoint| class. \pk{} uses instances of the Pint \pybox|Quantity| class to
store values with units, while Cantera expects a floating-point value in SI
units as input. Therefore, we use the built-in capabilities of Pint to convert
the units from those specified in the \ck{} file to SI units:
%
\begin{pythonbox}
T_initial = dp.temperature.to('K').magnitude
P_initial = dp.pressure.to('Pa').magnitude
X_initial = dp.get_cantera_mole_fraction()
